Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2025-04-29 manuscript.Rmd
minor edits requested by AFS

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
abstract: |
  Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and/or other woody perennials into agricultural land, thereby diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow (CF), silvopastures (SP), young agroforests (YA), home gardens (HG), and forest gardens (FG). Our results show that these traditional AFS can store large amounts of carbon, with the average being 155\ Mg\ C\ ha$^{-1}$. The AFS with the highest total carbon stocks (FG) stored an average of 213\ Mg\ C\ ha$^{-1}$. The average stocks stored by the other AFS were 108\ Mg\ C\ ha$^{-1}$ in CF, 158\ Mg\ C\ ha$^{-1}$ in SP, 134\ Mg\ C\ ha$^{-1}$ in YA, and 171\ Mg\ C\ ha$^{-1}$ in HG. Biomass carbon stocks varied substantially between AFS (which differed in tree cover), while soil carbon stocks were less variable between AFS but more site-dependent. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
  @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
keywords: "Traditional agroforestry, carbon stocks, soil, biomass"
editor_options: 
  markdown: 
    wrap: 72
---

@@

```{r setup, include=FALSE}
# Run with R version 4.3.2

packages_needed = c(
  "ggplot2",      # version 3.4.4
  "ggrepel",      # version 0.9.4
  "ggspatial",    # version 1.1.9
  "ggpubr",       # version 0.6.0
  "ggfortify",    # version 0.4.17
  "maps",         # version 3.4.2
  "rdryad",       # version 1.0.0
  "readxl",       # version 1.4.3
  "data.table",   # version 1.14.8
  "knitr",        # version 1.45
  "BIOMASS",      # version 2.1.11
  "terra",        # version 1.7-55
  "car",          # version 3.1-2
  "lmerTest",     # version 3.1-3
  "lme4",         # version 1.1-35.1
  "Matrix",       # version 1.6
  "performance",  # version 0.10.8
  "flextable"     # version 0.9.5
)

packages_to_install = packages_needed[!( packages_needed %in% rownames(installed.packages()))]

if (length(packages_to_install) > 0)
  install.packages(packages_to_install)

@@

opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, results = "hide")
```

@@

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation[@Albrecht2003; @Verchot2007; @Lasco2014; @Cardinael2021]. Agroforestry is the collective name for land-use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry practices thus allow for the diversification of agricultural production, making agricultural systems more resilient to climate change [@Duffy2021; @Terasaki2023]. Agroforestry practices have long been used in traditional agricultural systems for both environmental and economic benefits [@Viswanath2017].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in vegetation biomass and soil. This carbon storage capacity depends on the type of the AFS as well as environmental conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in the vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable. Although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e. planting trees with intercropping during the establishment phase) may decrease SOC stocks, possibly due to the initial soil disturbance caused by tree planting [@Feliciano2018]. Conversely, AFS such as alley cropping and home gardens have been shown to increase SOC stocks [@Oelbermann2004; @Shi2018], while other AFS such as silvopastures have shown mixed results [@Feliciano2018; @Shi2018]. Other factors such as climate, soil characteristics and previous land-use type can also influence SOC in AFS, which may explain the variability of the observations reported in the literature [@Nair2009; @Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS has only been evidenced to a depth of 10 cm [@Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009; @Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country and the livelihoods of its people. AFS have been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022; @Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022; @Cogne2024] and can be key elements for biodiversity conservation [@Perfecto2008; @Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks will improve understanding of the ecological importance of AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

@@

  ## crop to reduce memory usage
  box = c(124.5, 127.5, -9.2, -8.3) # Timor-Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  writeRaster(rst, filename = destfile, overwrite = TRUE) 
}

@@

```{r extract-env-data}
# extract elevation, precipitation and temperature for all plots
data_plot[, `:=`(
  elev = terra::extract(terra::rast("data/srtm_62_14.tif"), 
                        cbind(long, lat))[[1]], 
  temp = terra::extract(terra::rast("data/CHELSA_bio1_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  prec = terra::extract(terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  varp = terra::extract(terra::rast("data/CHELSA_bio15_1981-2010_V.2.1.tif"), 
                        cbind(long, lat))[[1]], 
  cwd = terra::extract(terra::rast("data/CWD.tif"), 
                       cbind(long, lat))[[1]]
)]

# extract elevation and rainfall for the 2 sites
site_env = data_plot[, as.list(expand.grid(seq(min(long), max(long), 0.001), 
                                           seq(min(lat), max(lat), 0.001))),
                     .(site)]

colnames(site_env) = c("site", "long", "lat")

@@

site_summary = site_env[, .(
  coord = paste0(abs(round(mean(lat),2)), "°S, ", round(mean(long),2), "°E"), 
  prec = round(mean(prec)), 
  elev = paste(round(quantile(elev, c(0.025, 0.975))/10)*10, collapse = " and ")),
  .(site)]
```

The research was carried out in the centre of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (`r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (masl). The second, the village of Osso-Luga in Samalari *suco* (`r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` masl (Figure 1). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment. The land use of these two villages is typical of the Baucau municipality. The five AFS described in this article [see detailed description in @Cogne2024] are found in the valleys and high plateaus, while non-AFS such as rainfed rice and horticultural plots are more likely to be found in the lowlands. Other non-agricultural land uses are also present, such as limestone quarrying. Some natural forest parks are also preserved around mountain peaks, such as Matebian and Mundo Perdido. Due to the risk of flooding, villages are rarely located near rivers, and are concentrated instead near paved roads and on hilltops. 
## Typology of agroforestry systems

@@

Within each sampling plot, all trees with a circumference at breast height (CBH) greater than or equal to 100 cm were identified taxonomically and their CBH and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the centre of the sampling plot (Figure S\@ref(fig:design)). Within these subplots, all trees with a CBH between 30 and 100 cm were taxonomically identified and their CBH and height were measured. Height measurements were taken with a dendrometer.

@@

# remove NA or <30 cm circumference values or < 100 cm outside subplots
data_tree <- subset(data_tree, 
                    !is.na(DIAM_130_CM) & 
                      DIAM_130_CM >= 30/pi &
                      (DIAM_130_CM >= 100/pi | grepl("Q", N_Q_ID)))

@@

# get carbon content 
GLOWCAD_files = dryad_download("10.5061/dryad.18931zcxk")
GLOWCAD = read.csv(GLOWCAD_files[[1]][3]) |> 
  subset(grepl("tree", growth.form)) |> 
  subset(tissue %in% c("stem", "root", "coarseroot")) |> 
  subset(biome == "tropical") 
data_ccontent <-

  # estimate mean AGC
  data_tree$mean <- computeAGB(
    D = data_tree$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    H = data_tree$HEIGHT
  ) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tree,
  AGBmonteCarlo(
    D = data_tree$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tree$HEIGHT, 
    errH = data_tree$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[GENUS %in% palm_genera, 
             HEIGHT := ifelse(
               iter == "mean", 
               HEIGHT,  ## height in m
               rnorm(length(HEIGHT), HEIGHT, HEIGHT*cvH)
             )]

## add WD data
data_wd_palms = unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc = merge(
  data_sim_agc,
  data_wd_palms,
  all.x = TRUE,
  by.x = "GENUS",
  by.y = "genus"
)
data_sim_agc[GENUS %in% palm_genera, 
             WD := ifelse(iter == "mean", 
                          meanWD, 
                          rnorm(length(meanWD), meanWD,sdWD))]
# palm biomass with Goodman et al 2013 equation that includes diameter, height
# and dry mass fraction; we used an average dmf of 0.37 which 
data_sim_agc[
  GENUS %in% palm_genera, 
  AGC := ((0.463 * DIAM_130_CM^2 * HEIGHT)^0.25 * 0.55512) ^ 4	* 0.4713 *1e-3]
```

We estimated tree-level AGC stocks from the tree CBH and height measurements using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. Palm biomass was estimated using the family level allometric equation of @Goodman2013 with DBH, stem height and dry mass fraction as predictors. As we didn't have species-specific values for dry mass fraction, we used the average value of 0.463 from @Goodman2013. Biomass was converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

data_sim_agc[, RS := exp(- 1.2312 - 0.0215 * DIAM_130_CM + 
                           0.0002 * DIAM_130_CM^2 - 0.0007 * cwd)]
data_sim_agc[, BGC := RS * AGC]
```

@@

## bootstrap AGC and BGC values in small plots (except for iter = mean)
data_sim = 
  merge(
    data_sim[!grepl("Q", N_Q_ID)], 
    rbind(
      data_sim[grepl("Q", N_Q_ID) & iter != "mean", .(
        valueSmall = ifelse(
          length(value) == 0,   ## some quadrats have zero trees: complete their values
          0, 
          mean(sample(c(value, rep(0, 5 - length(value))), replace = TRUE))
        )), .(N_PLOT, variable, iter)], 
      data_sim[grepl("Q", N_Q_ID) & iter == "mean", 
               .(valueSmall = sum(value / 5)), .(N_PLOT, variable, iter)]
    ),
    by = c("N_PLOT", "variable", "iter"), all = TRUE
  )

@@

## add zeros for plots with no tree above 10 cm DBH
# deal with non detectable values: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9216349/
# use uniform distribution between zero and limit of detectability
data_sim_empty = expand.grid(N_PLOT = setdiff(data_plot$N_PLOT, data_sim$N_PLOT), 
                             iter = unique(data_sim$iter), 
                             variable = c("AGC", "BGC"))|> 
  setDT() |> merge(data_plot[, c("N_PLOT", "ID_VILLAGE", "ID_AF")]) 
data_sim_empty$value = 0

@@

data_sim = rbind(data_sim[, colnames(data_sim_empty), with = FALSE], data_sim_empty)
```

@@

\begin{equation}
SOC_p = \sum_i Corg_{i,p} \cdot BDfine_{i,p}\cdot (1- \frac{Vcoarse_{i,p}}{Vcyl})\cdot (Surf \cdot \Delta d) \cdot 10^{-6}
\end{equation}

Where $ Corg_{i,p}$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger (depths 0-10 cm and 20-30 cm) and estimated at depth 10-20 cm (see details below). $BDfine_{i,p}$ is the bulk density of the fine fraction of the soil sampled with the cylinder at depth $i$ (in g cm$^{-3}$):

$$BDfine_{i,p} = \frac{mfine_{i,p}}{Vcyl – Vcoarse_{i,p}}$$
$mfine_{i,p}$ is the mass of the fine fraction (\< 2 mm) of the sampled soil; Vcyl is the volume of the sampling cylinder (250 cm$^3$); and $Vcoarse_{i,p}$ is the volume of the coarse fraction in the sampling cylinder. $Surf = 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm); 10^{-6}$ is the conversion factor from g to Mg.

@@

\begin{equation}
Corg_{10-20,p}=Corg_{0-10,p}\cdot\sqrt{\frac{Corg_{20-30,p}}{Corg_{0-10,p}}}
(\#eq:soc_depth)
\end{equation}

where $Corg_{a-b,p}$ is the average SOC content in plot $p$ between depths $a$ and $b$
(see rationale for equation \@ref(eq:soc_depth) in Supplemental Information).

@@

data_soil <- merge(data_soil_lab, data_soil_field, by = c("N_PLOT", "SAMPLE_DEPTH"), all=TRUE)
setDT(data_soil)
```

```{r add-missing-values}
data_soil[, Corg := `Corg\n`]
## add intermediate soil carbon in 10-20 cm depth under the assumption of an
# organic carbon content that decreases as a negative exponential of depth, we can deduce the carbon content at 10-20 cm with the following function: SOC[10-20] = SOC[0-10]*sqrt(SOC[20-30]/SOC[0-10])
data_soil = merge(
  data_soil, 
  data_soil[, .(
    Corg1020 = Corg[SAMPLE_DEPTH=="0-10"]*
      sqrt(Corg[SAMPLE_DEPTH=="20-30"]/Corg[SAMPLE_DEPTH=="0-10"])), 
    .(N_PLOT)])

data_soil[SAMPLE_DEPTH=="10-20", Corg := Corg1020]
```

@@

## error propagation using a Monte Carlo method -------------
data_sim_soil = data_soil[, .(mfine = c(Sample_DRY, rnorm(1000, Sample_DRY, Sample_DRY*0.1)), # 10% error on fine soil mass
                              Corg = c(Corg, rnorm(1000, Corg, Corg*0.12)), # 12% error on C content measurement
                              iter = c("mean", paste0("V", 1:1000))), 
                          .(N_PLOT, SAMPLE_DEPTH, h1, h0)]

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Camille Piponiot, Marguerite Cogné, Vincent Freycon, Alexis Thoumazeau, Marçal Gusmão, Régis Peltier"
abstract: |
  Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and/or other woody perennials into agricultural land, thereby diversifying landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste, namely cropping systems with fallow (CF), silvopastures (SP), young agroforests (YA), home gardens (HG), and forest gardens (FG). Our results show that these traditional AFS can store large amounts of carbon, with the average total carbon stocks (soil and biomass) being 155\ Mg\ C\ ha$^{-1}$. The AFS with the highest total carbon stocks (FG) stored an average of 213\ Mg\ C\ ha$^{-1}$. The average stocks stored by the other AFS were 108\ Mg\ C\ ha$^{-1}$ in CF, 158\ Mg\ C\ ha$^{-1}$ in SP, 134\ Mg\ C\ ha$^{-1}$ in YA, and 171\ Mg\ C\ ha$^{-1}$ in HG. Biomass carbon stocks varied substantially between AFS (which differed in tree cover), while soil carbon stocks were less variable between AFS but more site-dependent. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional AFS in Timor-Leste and their high carbon sequestration capacity. These results could provide an important baseline for the inclusion of AFS in Timor-Leste’s climate change mitigation strategy, and could serve as a reference for future AFS studies in different agro-climates of Timor-Leste.

journal: "Agroforestry systems"
date: ""
output:
  officedown::rdocx_document: 
    base_format: "bookdown::word_document2"
    reference_docx: misc/word-styles-ref-02.docx
bibliography: misc/references.bib
nocite: |
  @Nair1993, @Nair2021
csl: misc/springer-basic-author-date.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

**Keywords**: Vegetation biomass, soil organic carbon, farming systems, climate change mitigation, South East Asia.

```{r setup, include=FALSE}
# Run with R version 4.4.1
# package management is done with renv
packages_needed <- c(
  "ggplot2", "ggrepel",  "ggspatial",  "ggpubr",  "ggfortify",  "maps", 
  "rdryad", "readxl", "data.table", "knitr", "BIOMASS", "terra", "car", 
  "flextable"
)

@@

opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = "hide")
```

@@

Agroforestry has attracted attention in recent decades as a set of effective agricultural practices that could contribute to climate change mitigation and adaptation [@Albrecht2003; @Verchot2007; @Lasco2014; @Cardinael2021]. Agroforestry is the collective name for land-use systems and technologies where woody perennials (trees, shrubs, bamboos, creepers, etc.) are deliberately used on the same land management units as agricultural crops and/or animals in some form of spatial arrangement or temporal sequence [@Lundgren1982]. Agroforestry practices thus allow for the diversification of agricultural production, making agricultural systems more resilient to climate change [@Duffy2021; @Terasaki2023]. Agroforestry practices have long been used in traditional agricultural systems for both environmental and economic benefits [@Viswanath2017].

Agroforestry systems (AFS) have the potential to store large amounts of carbon in vegetation biomass and soil. This carbon storage capacity depends on the type of the AFS as well as environmental conditions [@Albrecht2003; @Feliciano2018]. Since trees can store large amounts of carbon, the carbon stored in the vegetation aboveground biomass of AFS is largely dependent on the number and size of trees in the system [@Ma2020]. However, the effect of AFS on soil organic carbon (SOC) is more variable. Although AFS appear to increase SOC stocks overall in tropical regions, some AFS such as woodlots (i.e., planting trees with intercropping during the establishment phase) may decrease SOC stocks, possibly due to the initial soil disturbance caused by tree planting [@Feliciano2018]. Conversely, AFS such as alley cropping and home gardens have been shown to increase SOC stocks [@Oelbermann2004; @Shi2018], while other AFS such as silvopastures have shown mixed results [@Feliciano2018; @Shi2018]. Other factors such as climate, soil characteristics and previous land-use type can also influence SOC in AFS, which may explain the variability of the observations reported in the literature [@Nair2009; @Feliciano2018; @Ma2020; @Martin2020]. Furthermore, the relationship between the amount of carbon stored in aboveground vegetation and soil in AFS has only been evidenced to a depth of 10 cm [@Ma2020].

In Timor-Leste, a large proportion of the population depends on agriculture for their livelihoods. However, population growth, increasing pressure on natural resources, and an increasingly variable climate with an unreliable rainfall regime are threatening food security [@WB2009; @Molyneux2012]. In this context, productive, resilient, and sustainable agriculture is critical for the country and the livelihoods of its people. AFS have been presented as a potential solution to many of the challenges facing Timorese agriculture [@Paudel2022; @Cogne2024]. Although trees already form a large part of traditional agricultural systems in Timor-Leste [@Paudel2022; @Cogne2024; @Gusmao2025] and can be key elements for biodiversity conservation [@Perfecto2008; @Torquebiau1992], they have received little attention and, to our knowledge, no study has quantified carbon storage in traditional Timorese AFS. A precise quantification of these carbon stocks will improve understanding of the ecological importance of AFS at both local and global scales, and help integrate them into national and international climate change mitigation initiatives.

@@

  ## crop to reduce memory usage
  box <- c(124.5, 127.5, -9.2, -8.3) # Timor-Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  writeRaster(rst, filename = destfile, overwrite = TRUE)
}

@@

```{r extract-env-data}
# extract elevation, precipitation and temperature for all plots
data_plot[, `:=`(
  elev = terra::extract(
    terra::rast("data/srtm_62_14.tif"),
    cbind(long, lat)
  )[[1]],
  temp = terra::extract(
    terra::rast("data/CHELSA_bio1_1981-2010_V.2.1.tif"),
    cbind(long, lat)
  )[[1]],
  prec = terra::extract(
    terra::rast("data/CHELSA_bio12_1981-2010_V.2.1.tif"),
    cbind(long, lat)
  )[[1]],
  varp = terra::extract(
    terra::rast("data/CHELSA_bio15_1981-2010_V.2.1.tif"),
    cbind(long, lat)
  )[[1]],
  cwd = terra::extract(
    terra::rast("data/CWD.tif"),
    cbind(long, lat)
  )[[1]]
)]

# extract elevation and rainfall for the 2 sites
site_env <- data_plot[
  , as.list(expand.grid(
    seq(min(long), max(long), 0.001),
    seq(min(lat), max(lat), 0.001)
  )),
  .(site)
]

colnames(site_env) <- c("site", "long", "lat")

@@

site_summary <- site_env[
  , .(
    coord = paste0(abs(round(mean(lat), 2)), "°S, ", round(mean(long), 2), "°E"),
    prec = round(mean(prec)),
    elev = paste(round(quantile(elev, c(0.025, 0.975)) / 10) * 10, collapse = " and ")
  ),
  .(site)
]
```

The research was carried out in the centre of the Baucau municipality, located on the north-eastern coast of Timor-Leste. Two communities, or *suco* (a group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases. The first, Gariuai *suco* (`r site_summary[site=="Gariuai", coord]`), is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging between `r site_summary[site=="Gariuai", elev]` m above sea level (masl). The second, the village of Osso-Luga in Samalari *suco* (`r site_summary[site=="Osso-Luga", coord]`), is located on the Matebian foothills based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso-Luga", prec]` mm and an altitude ranging between `r site_summary[site=="Osso-Luga", elev]` masl (Figure 1). The soils of the Baucau municipality are classified using the WRB soil classification [@Mantel2023] as Lithosols and Calcic Luvisols [@FAO2012]. The average soil properties differ between the two sites (Table S\@ref(tab:site-soil)); Gariuai has a clay loam texture whereas Osso-Luga has a clay texture. The two villages also have different histories. Gariuai has been inhabited for at least 50 years, while Osso-Luga was abandoned during the Indonesian occupation and was only repopulated after Timor-Leste gained independence in 2002. The AFS in Osso-Luga lay fallow during the period of abandonment. The land use of these two villages is typical of the Baucau municipality. The five AFS described in this article [see detailed description in @Cogne2024] are found in the valleys and high plateaus, while non-AFS such as rainfed rice and horticultural plots are more likely to be found in the lowlands. Other non-agricultural land uses are also present, such as limestone quarrying. Some natural forest parks are also preserved around mountain peaks, such as Matebian and Mundo Perdido. Due to the risk of flooding, villages are rarely located near rivers, and are concentrated instead near paved roads and on hilltops.

## Typology of agroforestry systems

@@

Within each sampling plot, all trees with a circumference at breast height greater than or equal to 100 cm were identified taxonomically and their circumference at breast height and height were measured. Five subplots of 10x10 m (0.01 ha) were established at each corner and in the centre of the sampling plot (Figure S\@ref(fig:design)). Within these subplots, all trees with a circumference at breast height between 30 and 100 cm were taxonomically identified and their circumference at breast height and height were measured. Height measurements were taken with a dendrometer.

@@

# remove NA or <30 cm circumference values or < 100 cm outside subplots
data_tree <- subset(
  data_tree,
  !is.na(DIAM_130_CM) &
    DIAM_130_CM >= 30 / pi &
    (DIAM_130_CM >= 100 / pi | grepl("Q", N_Q_ID))
)

@@

# get carbon content
GLOWCAD_files <- dryad_download("10.5061/dryad.18931zcxk")
GLOWCAD <- read.csv(GLOWCAD_files[[1]][3]) |>
  subset(grepl("tree", growth.form)) |>
  subset(tissue %in% c("stem", "root", "coarseroot")) |>
  subset(biome == "tropical")
data_ccontent <-
  # estimate mean AGC
  data_tree$mean <- computeAGB(
    D = data_tree$DIAM_130_CM,
    WD = data_wd$meanWD,
    H = data_tree$HEIGHT
  ) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tree,
  AGBmonteCarlo(
    D = data_tree$DIAM_130_CM,
    WD = data_wd$meanWD,
    errWD = data_wd$sdWD,
    H = data_tree$HEIGHT,
    errH = data_tree$HEIGHT * cvH,
    Dpropag = "chave2004",
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |>
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name = "AGC"
  )

@@

## height in cm, with a conversion factor between 0.8 and 1
data_sim_agc[
  GENUS %in% palm_genera,
  HEIGHT := ifelse(
    iter == "mean",
    HEIGHT, ## height in m
    rnorm(length(HEIGHT), HEIGHT, HEIGHT * cvH)
  )
]

## add WD data
data_wd_palms <- unique(data_wd[data_wd$genus %in% palm_genera, c("genus", "meanWD", "sdWD")])
data_sim_agc <- merge(
  data_sim_agc,
  data_wd_palms,
  all.x = TRUE,
  by.x = "GENUS",
  by.y = "genus"
)
data_sim_agc[
  GENUS %in% palm_genera,
  WD := ifelse(iter == "mean",
    meanWD,
    rnorm(length(meanWD), meanWD, sdWD)
  )
]
# palm biomass with Goodman et al 2013 equation that includes diameter, height
# and dry mass fraction; we used an average dmf of 0.37 which
data_sim_agc[
  GENUS %in% palm_genera,
  AGC := ((0.463 * DIAM_130_CM^2 * HEIGHT)^0.25 * 0.55512)^4 * 0.4713 * 1e-3
]
```

We estimated tree-level AGC stocks from the tree circumference at breast height and height measurements using the pantropical allometric equation from @Chave2014 and the R package BIOMASS [@Rejou-Mechain2017]. Each tree was assigned a mean and standard deviation of wood density at the species or genus level when at least one value was available from the Global Wood Density Database [@Zanne2009]. Other taxa were assigned a plot-level average wood density. Palm biomass was estimated using the family level allometric equation of @Goodman2013 with DBH, stem height and dry mass fraction as predictors. As we didn't have species-specific values for dry mass fraction, we used the average value of 0.463 from @Goodman2013. Biomass was converted into carbon with a factor of 0.4713 [@Thomas2012].

@@

data_sim_agc[, RS := exp(-1.2312 - 0.0215 * DIAM_130_CM +
  0.0002 * DIAM_130_CM^2 - 0.0007 * cwd)]
data_sim_agc[, BGC := RS * AGC]
```

@@

## bootstrap AGC and BGC values in small plots (except for iter = mean)
data_sim <-
  merge(
    data_sim[!grepl("Q", N_Q_ID)],
    rbind(
      data_sim[grepl("Q", N_Q_ID) & iter != "mean", .(
        valueSmall = ifelse(
          length(value) == 0, ## some quadrats have zero trees: complete their values
          0,
          mean(sample(c(value, rep(0, 5 - length(value))), replace = TRUE))
        )
      ), .(N_PLOT, variable, iter)],
      data_sim[
        grepl("Q", N_Q_ID) & iter == "mean",
        .(valueSmall = sum(value / 5)), .(N_PLOT, variable, iter)
      ]
    ),
    by = c("N_PLOT", "variable", "iter"), all = TRUE
  )

@@

## add zeros for plots with no tree above 10 cm DBH
# deal with non detectable values: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9216349/
# use uniform distribution between zero and limit of detectability
data_sim_empty <- expand.grid(
  N_PLOT = setdiff(data_plot$N_PLOT, data_sim$N_PLOT),
  iter = unique(data_sim$iter),
  variable = c("AGC", "BGC")
) |>
  setDT() |>
  merge(data_plot[, c("N_PLOT", "ID_VILLAGE", "ID_AF")])
data_sim_empty$value <- 0

@@

data_sim <- rbind(data_sim[, colnames(data_sim_empty), with = FALSE], data_sim_empty)
```

@@

\begin{equation}
SOC_p = \sum_i Corg_{i,p} \cdot BDfine_{i,p}\cdot (1- \frac{Vcoarse_{i,p}}{Vcyl})\cdot (Surf \cdot \Delta d) \cdot 10^{-6} (\#eq:socbd)
\end{equation}

Where $Corg_{i,p}$ is the SOC content at depth $i$ (either 0-10, 10-20 or 20-30 cm) as measured in the laboratory from the soil sampled by the auger (depths 0-10 cm and 20-30 cm) and estimated at depth 10-20 cm (see details below). $BDfine_{i,p}$ is the bulk density of the fine fraction of the soil sampled with the cylinder at depth $i$ (in g cm$^{-3}$):

$$BDfine_{i,p} = \frac{mfine_{i,p}}{Vcyl – Vcoarse_{i,p}}$$
$mfine_{i,p}$ is the mass of the fine fraction (\< 2 mm) of the sampled soil; $Vcyl$ is the volume of the sampling cylinder (250 cm$^3$); and $Vcoarse_{i,p}$ is the volume of the coarse fraction in the sampling cylinder. $Surf = 10^8$ is the surface conversion factor from cm$^2$ to ha; $\Delta d$ is the difference in sample depths (= 10 cm); 10$^{-6}$ is the conversion factor from g to Mg.

@@

\begin{equation}
Corg_{10-20,p}=Corg_{0-10,p}\cdot\sqrt{\frac{Corg_{20-30,p}}{Corg_{0-10,p}}}
(\#eq:socdepth)
\end{equation}

where $Corg_{a-b,p}$ is the average SOC content in plot $p$ between depths $a$ and $b$
(see rationale for equation \@ref(eq:socdepth) in Supplemental Information).

@@

data_soil <- merge(data_soil_lab, data_soil_field, by = c("N_PLOT", "SAMPLE_DEPTH"), all = TRUE)
setDT(data_soil)
```

```{r add-missing-values}
data_soil[, Corg := `Corg\n`]
## add intermediate soil carbon in 10-20 cm depth under the assumption of an
# organic carbon content that decreases as a negative exponential of depth, we can deduce the carbon content at 10-20 cm with the following function: SOC[10-20] = SOC[0-10]*sqrt(SOC[20-30]/SOC[0-10])
data_soil <- merge(
  data_soil,
  data_soil[
    , .(
      Corg1020 = Corg[SAMPLE_DEPTH == "0-10"] *
        sqrt(Corg[SAMPLE_DEPTH == "20-30"] / Corg[SAMPLE_DEPTH == "0-10"])
    ),
    .(N_PLOT)
  ]
)

data_soil[SAMPLE_DEPTH == "10-20", Corg := Corg1020]
```

@@

## error propagation using a Monte Carlo method -------------
data_sim_soil <- data_soil[
  , .(
    mfine = c(Sample_DRY, rnorm(1000, Sample_DRY, Sample_DRY * 0.1)), # 10% error on fine soil mass
    Corg = c(Corg, rnorm(1000, Corg, Corg * 0.12)), # 12% error on C content measurement
    iter = c("mean", paste0("V", 1:1000))
  ),
  .(N_PLOT, SAMPLE_DEPTH, h1, h0)
]

data_sim_soil[
  , Cstock := ## in Mg/ha
    (h1 - h0) * 1e8 * ## equivalent volume of a 1 ha soil sample, in cm3/ha
      Corg / 100 * ## carbon organic content in fine soil (unitless)
      (mfine * 1e-6) * ## mass of the fine fraction of the soil (Mg)
      (1 / 250) ## 1/volume of the cylinder, in cm-3
]
